Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest

### 2019-12-10 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Start on JoE revisions

\documentclass[letterpaper,12pt]{article}
\usepackage{jecol}

\title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\citep{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \citep{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \citep{Rothermel-1972, Nelson-2001}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons, and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \citep{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may strongly influence
relatively flammability across species at any particular time following
wetting. Such differential moisture content across time, may contribute to the
often high spatial heterogeneity of surface fires
\citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

Litter packing, measured as litter-bed density or packing ratio, controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability as a consequence of oxygen limitation. Leaf
litter density is driven by leaf size and shape
\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established. However, how moisture dynamics modulate species specific
flammability is poorly known and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns, but
current estimates of litter flammability in mixed conifer forests are based
largely on burning dry litter.

Fuel moisture content through time is controlled by two main processes,
absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
Absorptive capacity represents the potential maximum fuel moisture and is
evaluated by the moisture content of litter when saturated. After saturation,
the differences in moisture content indicate different abilities to retain
moisture. Litter particle geometry may have direct effects on maximum moisture
absorption through area-mass relationships and we also expect effects of leaf
morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
rates should also be influenced by leaf and litter particle traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore has two pathways of effect on
fire behavior: directly by its effect on oxygen availability for combustion,
and indirectly by its effect on dry-down rate, and both these pathways are
affected by leaf litter traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
moisture retainability and should favor drying, thus increasing the probability
of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects are synergistic in dry litter fuels where
the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. It is possible
that similar non-additive effects may influence litter moisture content,
especially if different leaf traits influence absorptive capacity and drying
rate. There is some evidence for non-additive flammability in moist fuels with
one study finding increasing non-additivity with moisture
\citep{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits. We then examined the influence
of moisture on the flammability across these species. Experimental burn trials
of leaf litter at various moisture levels determined how time since wetting
influenced two axes of flammability characterized by flame spread rate and
total fuel consumed. Previous work described the flammability of these species
in monocultures and in mixtures for oven-dry litter
\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non-additivity was maintained in
mixtures across varying levels of fuel moisture. The burn trials were also
performed with multi-species litter mixtures to evaluate if species contribute
equally to litter moisture dynamics or if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on
reported in previous studies occur under ecologically-relevant moisture levels
and whether fuel moisture effects strengthen or moderate such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in
mixed-conifer forest. We chose eight dominant tree species that are
representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf.,
\emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the
previous year (mostly undecomposed). Leaf litter was collected in the summer
(mid-June to mid-July) in 2012 from 21 separate sites across the study area. We
collected from a minimum of four populations per species with each population
9--32\,km distant from its neighbors. An exception to this was \emph{A.
  magnifica} which is restricted to the higher elevations and could only be
collected from one area due to designated wilderness boundaries and logistical
constraints. Within that area, \emph{A. magnifica} litter was collected from
over 10 different individuals. For the remaining species, the collection
involved 2--4 individual trees at each site which were separated from one
another by at least 10\,m. The litter was collected 2\,m away from the trunk to
obtain a more uniform sample of leaf and fine branchlet litter, because bark
and heavier branches tend to fall closer to the tree trunk. All litter was air
dried to $< 5$\% fuel moisture by dry weight for weighing and dividing into
moisture trials. The moisture manipulations and subsequent burning trials were
conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of 1\,cm aluminum screen. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
completely randomized design. During the dry-down process three subsamples of
5--10\,g were taken at 24 hour intervals and weighed. These subsamples were
then oven dried for 24 hours at 100\,C and re-weighed to assess moisture
content. This was repeated until the final sample from a basket was less than
10\% moisture.

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
measured on freshly fallen litter by sampling 5 individuals per species. This
measurement included petiole and for the scale-leaved species included the full
excised branchlet. Therefore, this is not directly comparable to typical specific leaf
area measurements on fresh leaves. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition
will not occur above 100\% fuel moisture. To expand our data on the dry end of
this gradient, we supplemented these data with data from an earlier experiment
on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
added five replicates per species at an oven dried moisture content of $<$ 5\%.
We assigned these replicates a time since wetting value of 144 hours equivalent
to the longest time since drying for the wetted samples. Flammability was
assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
leaf litter was placed and the table gently shaken to allow settling. Burning
trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
was 150\,g per species. This design allows fire to reach constant flame spread
rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite
after 30 seconds, no further attempt was made and if three samples of a species
at a particular moisture level failed to ignite, no further attempts were made
on that species/moisture combination and the replicate was allowed to continue
drying. A timer was used to record total time to ignition and duration of
flaming combustion. The flammability trials followed the methodology detailed
in de Magalhães and Schwilk \citep{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and mass loss were recorded, but flame spread rate and flame height
were recorded as zero. Our thermocouple system failed or data was lost for two
days of burning trials, therefore we have omitted those data from this
analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
captures a portion of the second major axis of flammability variation (although
not as well as integrating temperature over time does). Our analysis,
therefore, focused on these two relatively orthogonal measures of flammability
to avoid examining multiple redundant dependent variables.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficients we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R 3.6.1
\citep{RCoreTeam-2019}. We fit models with R packages `lme4'
\citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
Luke \citeyear{Luke-2017} as producing acceptable Type I error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

@@

Across these eight species specific leaf area had a positive effect on maximum
water retention (SI Fig 4, linear model $P$ = 0.003) and litter
bulk density had a negative effect on desiccation rate (SI Fig 5,
linear model $P$ = 0.029). ANOVA results are shown in Tables
\ref{tab:mc_di_anova}.

@@

  \begin{tabular}{llrrrr}
\toprule
Model set & model & Log Lik. & df & Deviance & $P$ \\ 
\midrule
  Effect of moisture &   \~{} 1 & 262.12 &  &  &  \\ 
   &  \~{} taxon & 268.73 & 3 & 13.23 & 0.0042 \\ 
   & \~{} taxon + mc & 289.57 & 1 & 41.68 & $<$ 0.0001 \\ 
   & \~{} taxon + mc + taxon:mc & 293.87 & 3 & 8.60 & 0.0351 \\ 
\midrule
Effect of time &  consum \~{} 1 & 140.08 &  &  &  \\ 
 & \~{} taxon & 155.84 & 4.00 & 31.51 & $<$ 0.001 \\ 
 & \~{} taxon + time & 209.84 & 1.00 & 108.01 & $<$ 0.001 \\ 
 & \~{} hour + taxon + taxon:time & 216.96 & 4.00 & 14.23 & 0.01 \\ 
    \bottomrule
    \end{tabular}
\end{table}

As a consequence of the relationships shown above, time since wetting had a
significant effect on spread rate ($P < 0.0001$) and the slope of this effect
varied by taxonomic group (Fig. \ref{fig:spread_time}, Table
\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much sooner
after wetting than did the other taxa, none of which had measurable spread
rates until burning was attempted 72 hours after wetting (note that trials that
failed to even ignite smoldering combustion are omitted completely from the
model and figure). At that point, the \emph{Abies}, \emph{Calocedrus}, and
\emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
spread rates were slow. \emph{Quercus} had the steepest response to time since
wetting (Fig. \ref{fig:spread_time}, SI Table 6)).

@@

Mixture dry down rates and maximum water retention showed some relationship to
the dry down curves for individual constituent species (see SI Fig. 6). For
example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
Fig. 6) has the lowest initial retention. Initial water retention and rates
differed across mixtures (linear model based on log of moisture content, SI Tables 8 and 9).

@@

There was evidence for non-additive effects in drying rate. Mixtures tended to
have lower water content than predicted from the mean of the constituent
species after wetting, but had higher water content than expected after long
drying (Wald test shows slope of mixed linear model illustrated in Fig.
\ref{fig:mixture_obs_pred_mc} less than 1, $P$ < 0.0001).

@@

\subsection*{Species-specific moisture dynamics flammability}

>>>>

%% Revision 2019-12-10
\documentclass[letterpaper,12pt]{article}
\usepackage{jecol}

\title{Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest}

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Moisture acts as a heat-sink resulting in less energy
available for propagation, and the resulting water vapor will both cool the
flaming front and dilute flammable gases \citep{Albini-1976, Shafizadeh-1977}.
Environmental conditions such as temperature, relative humidity, and wind drive
the moisture content of the fuel available to burn
\citep{Kreye+Varner+etal-2018}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. In particular, fine dead fuels respond very readily to environmental
changes \citep{Nelson-2001}. Fuel moisture influences flammability and fire
behavior through effects on ignition, spread rate, flame height, fuel
consumption, and heat release, due to the high specific heat content of water
\citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels varies with
climate, with seasons, and with short term weather within a season. Therefore,
the rate of drying and flammability as a function of moisture content are key
parameters that influence fuel flammability throughout the fire season
\citep{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however.  TODO REWRITE: In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may influence
relatively flammability across species and over time following
wetting. Such differential moisture content across time may contribute to the
often high spatial heterogeneity of surface fires
\citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

Litter packing, measured as litter-bed density or packing ratio, controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability as a consequence of oxygen limitation. Leaf
litter density is driven by leaf size and shape
\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established. However, how moisture dynamics modulate species specific
flammability is poorly known and could be driven by by leaf traits, emergent
litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual species traits have been recorded for western US forests as well: in mixed conifer forest in California, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns, but
current estimates of litter flammability in mixed conifer forests are based
largely on burning dry litter.

Fuel moisture content through time is controlled by two main processes,
absorption and retention \citep{Kreye+Kobziar+etal-2013}.
Absorptive capacity represents the potential maximum fuel moisture and is
evaluated by the moisture content of litter when saturated. Differences in moisture content over time since saturation can indicate different abilities to retain
moisture. Litter particle geometry may have direct effects on maximum moisture
absorption through area-mass relationships and we also expect effects of leaf
morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption, mmoisture retention should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore effects fire behavior in two ways: directly by its effect on oxygen availability for combustion,
and indirectly by its effect on dry-down rate. Both these pathways are
affected by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
moisture retainability should favor drying, thus increasing the probability
of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer [CITATIONS]  . Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects.

TODO

Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects are often synergistic in dry litter fuels where
the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. TODO cite negative flammability. Zhao et al 20198]

It is possible
that similar non-additive effects may influence litter moisture content,

% especially if different leaf traits influence absorptive capacity and drying
% rate.

There is some evidence for non-additive flammability in moist fuels with
one study finding increasing non-additivity with moisture
\citep{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined the influence
of moisture on the flammability across these species. Experimental burn trials
of leaf litter at various moisture levels determined how time since wetting
influenced two axes of flammability characterized by flame spread rate and
total fuel consumed. Previous work described the flammability of these species
in monocultures and in mixtures for oven-dry litter
\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non-additivity was maintained in
mixtures of varying levels of fuel moisture. The burn trials also
performed with multi-species litter mixtures to evaluate if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on
reported in previous studies occur under [TODO CHECK} ecologically-relevant moisture levels,
and whether fuel moisture effects strengthen or moderate such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National
Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
Buchholz. In this study, we consider litter the top layer of leaves and small
twigs less than 0.625\,cm diameter (= 1-hour fuel) that have fallen in the
previous year (mostly undecomposed or lightly decomposed). Leaf litter was collected in the summer
(mid-June to mid-July) in 2012 from 21 separate sites across the study area. We
collected from a minimum of four populations per species with each population
9--32\,km distant from its neighbors. An exception to this was \emph{A.
  magnifica} which is restricted to the higher elevations and could only be
collected from one area due to designated wilderness boundaries and logistical
constraints. Within that area, \emph{A. magnifica} litter was collected from
over 10 different individuals. For the remaining species, the collection
involved 2--4 individual trees at each site which were separated from one
another by at least 10\,m. The litter was collected 2\,m away from the trunk to
obtain a more uniform sample of leaf and fine branchlet litter, because bark
and heavier branches tend to fall closer to the tree trunk. The top 2.5-5\,cm
of litter were collected by raking and shoveling into large bags. All litter
was air dried to $< 5$\% fuel moisture by dry weight, then weighed and divided into moisture trials. The collected litter was dried in large batches, then
stored in sealed plastic bags inside cardboard boxes in a lab storage space. We
minimized handling and moving the bagged litter to avoid mechanical damage and
changes to the particle size distributions. The moisture manipulations and
subsequent burning trials were conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of aluminum mesh with 0.52\,cm square openings. Samples were
450\,g dry weight litter per trial and litter depth was standardized to 10\,cm
by changing the horizontal dimensions of the basket. Weight was measured by
using a balance sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus,
OH). Samples were immersed in plastic storage bins filled with water for 24
hours to fully saturate them. After saturation, the baskets were removed from
the water, allowed to drain for 3 minutes, and samples weighed to measure
maximum moisture content. The samples were then allowed to dry in a controlled
environment kept at $21^\circ$C temperature and 30\% relative humidity, assessed via
an iButton sensor (Maxim Integrated) placed over the samples. Litter baskets
were arranged in a completely randomized design. During the dry-down process
three subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to assess
moisture content. This was repeated until the final sample from a basket was
less than 10\% moisture.

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption (specific
leaf area), and one litter trait known to directly affect flammability and
likely to influence drying rate and litter packing density (litter bulk
density). [TODO check]  Specific leaf area was measured on freshly fallen litter by sampling
5 individuals per species. This measurement included petiole and for the
scale-leaved species included the full excised branchlet. Therefore, this is
not directly comparable to typical specific leaf area measurements on fresh
leaves. Litter bulk density was obtained in this experiment for each basket by
dividing the dry mass (450\,g) by the estimated litter volume (basket width x
length x litter depth).

@@

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition
will not occur above 100\% fuel moisture. To expand our data on the dry end of
this gradient, we supplemented these data with data from an earlier experiment
on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
added five replicates per species at an oven dried moisture content of $<$ 5\%.
We assigned these replicates a time since wetting value of 144 hours equivalent
to the longest time since drying for the wetted samples. Flammability was
assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
leaf litter was placed and the table gently shaken to allow settling. Burning
trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
was 150\,g per species. [TODO give depths] This design allows fire to reach
constant flame spread rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal
rulers equally spaced along the apparatus allowed visual determination of
maximum flame height. Litter was ignited with a propane torch; maximum
adiabatic flame temperature of 1899$^\circ$C,
\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
and allowed to burn until extinction. If a sample failed to ignite after 30
seconds, no further attempt was made and if three samples of a species at a
particular moisture level failed to ignite, no further attempts were made on
that species/moisture combination and the replicate was allowed to continue
drying until the next scheduled checkpoint. A timer was used to record total
time to ignition and duration of flaming combustion. The flammability trials
followed the methodology detailed in 
\citet{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and fuel consumption were recorded, but flame spread rate and flame height
were recorded as zero. Our thermocouple system failed or data was lost for two
days of burning trials, therefore we have omitted those data from this
analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018} and percent fuel consumed
captures a portion of the second major axis of flammability variation. Our analysis,
therefore, focused on these two relatively orthogonal measures of flammability
to avoid examining multiple redundant dependent variables.

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficients we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R 3.6.1
\citep{RCoreTeam-2019}. We fit models with R packages `lme4'
\citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
Luke \citeyear{Luke-2017} as producing acceptable Type I error rates. We tested if
one litter particle trait (specific leaf area) and one litter bed trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

@@

Across these eight species specific leaf area had a positive association with maximum
water retention (SI Fig 4, linear model $P$ = 0.003) and litter
bulk density had a negative effect on desiccation rate (SI Fig 5,
linear model $P$ = 0.029). ANOVA results are shown in Tables
\ref{tab:mc_di_anova}.

@@

  \begin{tabular}{llrrrr}
\toprule
Model set & model & Log Lik. & df & Deviance & $P$ \\ 
\midrule
  Effect of moisture &   \~{} 1 & 262.12 &  &  &  \\ 
   &  \~{} taxon & 268.73 & 3 & 13.23 & 0.0042 \\ 
   & \~{} taxon + mc & 289.57 & 1 & 41.68 & $<$ 0.0001 \\ 
   & \~{} taxon + mc + taxon:mc & 293.87 & 3 & 8.60 & 0.0351 \\ 
\midrule
Effect of time &  consum \~{} 1 & 140.08 &  &  &  \\ 
 & \~{} taxon & 155.84 & 4 & 31.51 & $<$ 0.001 \\ 
 & \~{} taxon + time & 209.84 & 1 & 108.01 & $<$ 0.001 \\ 
 & \~{} hour + taxon + taxon:time & 216.96 & 4 & 14.23 & 0.01 \\ 
    \bottomrule
    \end{tabular}
\end{table}

As a consequence of the relationships shown above, time since wetting had a
significant effect on spread rate ($P < 0.0001$) and the slope of this effect
varied by taxonomic group (Fig. \ref{fig:spread_time}, Table
\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much sooner
after wetting than did the other taxa, none of which had measurable spread
rates until burning was attempted 72 hours after wetting (note that trials that
failed to even ignite smoldering combustion are omitted completely from the
model and figure). At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
\emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
spread rates were slow. \emph{Quercus} had the steepest response to time since
wetting (Fig. \ref{fig:spread_time}, SI Table 6)).

@@

Mixture dry down rates and maximum water retention showed some relationship to
the dry down curves for individual constituent species (see SI Fig. 6). For
example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
Fig. 6) has the lowest initial retention. Initial water retention and drying rates
differed across mixtures (linear model based on log of moisture content, SI Tables 8 and 9).

@@

There was evidence for non-additive effects in drying rate. Mixtures tended to
have lower water content than predicted from the mean of the constituent
species after wetting, but had higher water content than expected after long
drying (Wald test shows slope of mixed linear model illustrated in Fig.
\ref{fig:mixture_obs_pred_mc} less than 1, $P < 0.0001$).

@@

\subsection*{Genus-specific moisture dynamics and flammability}

The genera in this study displayed different patterns of both moisture
absorption and desorption, associated with different litter traits. In
this sample of species, those that have a greater absorptive capacity are also
among the most flammable in terms of spread rate when dry. This highlights the
importance of fuel moisture in modulating flammability. One shortcoming of past
work in this system \citep{Magalhaes+Schwilk-2012} was that flammability trials
tested only dry fuels. We therefore expected that rankings of species
flammability might change under more [TODO FIX] realistic moisture levels. For example,
one presumption of ours prior to this experiment was that the high flammability
previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012} would
disappear under higher fuel moisture contents. This prediction was partly
supported: \emph{Q. kelloggii} fails to ignite until after 72 hours of drying
in this study (Figs. \ref{fig:spread_time} and \ref{fig:consume_time}).
However, high moisture content in this oak was unable to suppress ignition of
mixtures to which it contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
\ref{fig:mixture_obs_pred_consume}).
